---
title: "Exploratory Analysis of Spiritual Vocabulary"
author: "Alexandre PSI"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

## Introduction

This notebook provides an interactive exploration of the **Global Semantic Structure of Spirituality** datasets. We examine the distribution of spiritual terms across categories, traditions, and frequency bands, then visualise their semantic network and cluster structure.

---

## 1. Setup and Data Loading

```{r load-packages}
library(tidyverse)
library(readr)
library(ggplot2)
library(igraph)
library(ggraph)
library(cluster)
library(factoextra)
```

```{r load-data}
# Set working directory to project root before knitting, or use here::here()
spiritual_words <- read_csv("../data/spiritual_words.csv")
extended_words  <- read_csv("../data/extended_spiritual_words.csv")

cat("Core dataset rows:", nrow(spiritual_words), "\n")
cat("Extended dataset rows:", nrow(extended_words), "\n")
```

---

## 2. Exploratory Data Analysis

### 2.1 Dataset Overview

```{r glimpse}
glimpse(extended_words)
```

### 2.2 Word Counts by Category

```{r category-counts}
extended_words %>%
  count(category, sort = TRUE) %>%
  knitr::kable(caption = "Words per Category (Extended Dataset)")
```

### 2.3 Word Counts by Tradition

```{r tradition-counts}
extended_words %>%
  count(tradition, sort = TRUE) %>%
  knitr::kable(caption = "Words per Tradition")
```

### 2.4 Frequency Distribution

```{r freq-distribution}
ggplot(extended_words, aes(x = frequency, fill = category)) +
  geom_histogram(bins = 20, colour = "white") +
  facet_wrap(~category, scales = "free_y") +
  labs(
    title = "Frequency Distribution by Category",
    x     = "Relative Frequency",
    y     = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### 2.5 Top Words by Frequency

```{r top-words}
extended_words %>%
  arrange(desc(frequency)) %>%
  slice_head(n = 20) %>%
  ggplot(aes(x = reorder(word, frequency), y = frequency, fill = category)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 20 Spiritual Terms by Frequency",
    x     = "Word",
    y     = "Relative Frequency",
    fill  = "Category"
  ) +
  theme_minimal()
```

---

## 3. Semantic Network

```{r build-network}
words <- extended_words %>%
  distinct(word, .keep_all = TRUE)

n          <- nrow(words)
adj_matrix <- matrix(0, n, n, dimnames = list(words$word, words$word))

for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    if (i != j &&
        (words$category[i] == words$category[j] ||
         words$tradition[i] == words$tradition[j])) {
      adj_matrix[i, j] <- words$frequency[i] * words$frequency[j]
    }
  }
}

g <- graph_from_adjacency_matrix(adj_matrix, mode = "undirected",
                                  weighted = TRUE, diag = FALSE)
V(g)$category  <- words$category[match(V(g)$name, words$word)]
V(g)$frequency <- words$frequency[match(V(g)$name, words$word)]

set.seed(42)
V(g)$community <- membership(cluster_louvain(g))
```

```{r plot-network, fig.width=11, fig.height=9}
set.seed(42)
ggraph(g, layout = "fr") +
  geom_edge_link(aes(alpha = weight), colour = "grey70", show.legend = FALSE) +
  geom_node_point(aes(size = frequency, colour = category)) +
  geom_node_text(aes(label = name), repel = TRUE, size = 2.8, max.overlaps = 20) +
  scale_size_continuous(range = c(3, 9)) +
  labs(
    title  = "Semantic Network of Spiritual Vocabulary",
    colour = "Category",
    size   = "Frequency"
  ) +
  theme_graph()
```

---

## 4. Clustering Analysis

```{r clustering}
words_clean <- extended_words %>%
  distinct(word, .keep_all = TRUE) %>%
  mutate(
    category_num  = as.numeric(factor(category)),
    tradition_num = as.numeric(factor(tradition))
  )

features <- words_clean %>%
  select(frequency, category_num, tradition_num) %>%
  scale()
rownames(features) <- words_clean$word

set.seed(42)
km <- kmeans(features, centers = 5, nstart = 25)
words_clean$cluster <- factor(km$cluster)
```

```{r cluster-plot}
fviz_cluster(km, data = features, geom = "point",
             ellipse.type = "convex", palette = "jco",
             ggtheme = theme_minimal()) +
  labs(title = "K-Means Clustering of Spiritual Terms (k = 5)")
```

```{r cluster-composition}
words_clean %>%
  count(cluster, category) %>%
  ggplot(aes(x = cluster, y = n, fill = category)) +
  geom_col(position = "stack") +
  labs(
    title = "Cluster Composition by Category",
    x     = "Cluster",
    y     = "Number of Words",
    fill  = "Category"
  ) +
  theme_minimal()
```

---

## 5. Summary

This exploratory analysis reveals that the semantic field of spirituality is organised around several coherent conceptual clusters: **divine concepts**, **practices**, **inner states**, **virtues**, and **beliefs**. The semantic network confirms the central role of high-frequency terms such as *God*, *faith*, *love*, and *compassion* as hubs that bridge multiple traditions.

Future work should extend this analysis to:

1. Larger, cross-lingual corpora of spiritual texts.
2. Dynamic network analysis to trace semantic change over time.
3. Comparison of network structures across different religious traditions.

---

*This notebook was generated as part of the Global Semantic Structure of Spirituality project.*
